Better fusion for filters

### 2014-03-21 2014betterfusionforfilters/section/04-ILP.tex
ok

@@

\subsubsection{Satisfiability}
For any program $p$, there exists a trivial clustering with no fusion at all.
We can use this as the variable assignment of $ilp(p)$.
For each pair of nodes $m,n \in p$, $x_{mn} = 1$ - no fusion is possible.
For the $\pi$ variables, we must find a topographical ordering of the nodes in $p$, which is simple since we are assured it is a dag.

@@

\subsubsection{Legality}
For any program $p$ and variable asignment $v$, if $v$ satisfies the constraints for $ilp(p)$, the clustering denoted by $x_ij$ in $v$ is legal.

For a clustering to be legal, it must satisfy three constraints:
\begin{description}
\item[Acyclic]
after merging nodes of same cluster together, the resulting graph must be a dag
\item[Precedence preserving]
if there is an edge between two nodes $i$ and $j$, and they are not merged together, then we require $\pi_j > \pi_i$
\item[Type constraint]
if two nodes $i$ and $j$ are in the same cluster, then $\tau_i = \tau_j$, or if $\tau_i$ is a subtype of $\tau_j$ (or $\tau_j$ is a subtype of $\tau_i$), then the \emph{generator} for $\tau_i$ (or $\tau_j$) must also be in the same cluster as $i$ and $j$.
    \\
    \TODO This is wrong.
\end{description}

>>>>

\subsection{Function $\to$ graph}
Converting a function in \emph{combinator normal form} to a graph --- in fact, a DAG --- is quite simple.
Each binding becomes a node in the graph.
When a binding references other arrays or scalars, there is an edge between those two nodes.
Edges may be either \emph{fusible} or \emph{fusion-preventing}.
Fusion-preventing edges mean that the entire input node must completely finish its execution before the output node can start executing.
For example, @fold@s consume their entire input array before producing a single result, so any references to folds must be fusion-preventing.
Conversely, @map@s produce output data for every input element, so may be fused.

The @gather@ operation is interesting; it takes an indices array and a data array, and for each element in indices returns that element in data.
Thus, @gather@ requires random-access in its data array, and is not fusible, but consumes indices linearly and may fuse.

\subsection{Graph $\to$ ILP}

ILP variables:
\begin{tabbing}
M   \= M \= MMMMMMM \= MM \= \kill
$x$ \> @::@  \> $node \times node$ \> $\to$ \> $\mathbb{B}$ \\
$\pi$ \> @::@  \> $node$ \> $\to$ \> $\mathbb{N}$ \\
\end{tabbing}

Talk about how $x_{ij} = 0$ if and only if $i$ and $j$ are fused together, otherwise $x_{ij} = 1$. This may seem reversed, but makes certain constraints easier to formulate.

Use the lemma:
a graph $G$ is acyclic if and only if there exists some mapping $\pi :: node \to \mathbb{N}$ such that for every edge in $G$ from $i$ to $j$, $\pi_i < \pi_j$.

\begin{tabbing}
MMMMM   \= MMMMMMMMMMM \= \kill
Minimise   \> $\Sigma_{(i,j) \in E} W_{ij} x_{ij}$       \> ($W_{ij}$ is constant weight)      \\
Subject to \> $-N x_{ij} \le \pi_j - \pi_i \le N x_{ij}$ \> (no edge from i to j)            \\
           \> $              \pi_j - \pi_i \le N x_{ij}$ \> (an edge from i to j)            \\
\end{tabbing}

\subsection{Better than stream fusion}
Abdallah suggests something like:
\[
\forall g.\ \exists c.\ ilp(g, c) \le_c streamfusion(g)
\]
ie, for any graph there exists a criterion to optimise against, where we are better than or equal to stream fusion.

@@

\subsubsection{Satisfiability}
For any program $p$, there exists a trivial clustering with no fusion at all.
We can use this as the variable assignment of $ilp(p)$.
For each pair of nodes $m,n \in p$, $x_{mn} = 1$ --- no fusion is possible.
For the $\pi$ variables, we must find a topographical ordering of the nodes in $p$, which is simple since we are assured it is a dag.

@@

\subsubsection{Soundness}
For any program $p$ and variable asignment $v$, if $v$ satisfies the constraints for $ilp(p)$, the clustering denoted by $x_{ij}$ in $v$ is legal.

For a clustering to be legal, it must satisfy three constraints:
\begin{description}
\item[Acyclic]
after merging nodes of same cluster together, the resulting graph must be a dag
\item[Precedence preserving]
if there is an edge between two nodes $i$ and $j$, and they are not merged together, then we require $\pi_j > \pi_i$
\item[Type constraint]
if two nodes $i$ and $j$ are in the same cluster, then $\tau_i = \tau_j$, or if $\tau_i$ is a subtype of $\tau_j$ (or $\tau_j$ is a subtype of $\tau_i$), then the \emph{generator} for $\tau_i$ (or $\tau_j$) must also be in the same cluster as $i$ and $j$.
    \\
    \TODO Actually, let us say $x_{ij} = 0 \implies check_{ij}$

\end{description}
where
\begin{tabbing}
MMMMM      \= M \= MMMMMMM \= MM \= \kill
$check$ \> @::@  \> $array \times array$ \> $\to$ \> $\mathbb{B}$ \\
MMMMM      \= M \= MMMMMM \= MM \= \kill
$check(i, j)$     \> $|$ \> $tau_i = tau_j$ \> $=$ \> $x_{i,j} = 0$                        \\
$check(i, j)$     \> $|$ \> $i' \in gen(i) $ \> $=$ \> $x_{i',j} = 0 \wedge x_{i,i'} = 0 \wedge check(i', j)$                        \\
$check(i, j)$     \> $|$ \> $j' \in gen(j) $ \> $=$ \> $x_{i,j'} = 0 \wedge x_{j,j'} = 0 \wedge check(i, j')$                        \\
$check(i, j)$     \> $|$ \> $tau_i \not= tau_j$\> $=$ \> $\bot$
\end{tabbing}

### 2014-03-21 2014betterfusionforfilters/section/05-Related.tex
ok

\subsection{Haskell short-cut fusion}
Existing fusion systems for Haskell such as stream fusion\cite{coutts2007streamfusion}, tend to cleverly reuse compiler optimisations such as inlining and rewrite rules, to fuse combinators without having to modify the compiler itself.
This approach has the advantage of simplicity, but is inherently limited in the amount of fusion it can perform.

@@

Talk about the simple formulation by Darte\cite{darte2002new}.
Has an integer variable for each node, denoting the number of the cluster it's in.
Also includes a binary variable for each node, which is whether the node is fused with all its successors - in which case no array would be required, and an integer variable which is the maximum of all cluster numbers.
The objective function is to maximise the number of nodes that are completely fused, requiring no arrays, and minimise the maximum cluster number, which in turn minimises the number of clusters.
It doesn't require many constraints and is easy to implement, but doesn't work as well when there are loops of different sizes.
As loops of different sizes cannot be fused together, a simple method is to introduce an ordering on the sizes, and then extract loops of the same cluster number in order of size.
The problem here is that the objective function uses the maximum cluster number to minimise the number of loops, but this alone is no longer sufficient when there are multiple sizes. Explain why.

The formulation by Megiddo\cite{megiddo1997optimal} supports different sized loops, and is therefore more relevant for our purposes.
For every pair of nodes $i,j$ in the LDG, a variable $x_{ij}$ is created, which denotes whether $i$ and $j$ are fused together.
Slightly awkwardly, but for simplicity of other constraints, $x_{ij} = 0$ if the two nodes are fused together.
If there is a fusion preventing edge between $i$ and $j$, then $x_{ij}$ is constrained to be $1$ - that is, no fusion is possible.
This alone is not enough to guarantee a valid clustering.
To constrain the solution to acyclic and precedence preserving clusterings, a variable $\pi_i$ is added for each node $i$.
Constraints are added that require two nodes $i,j$ to have $\pi_i = \pi_j$ if $x_{ij} = 0$, and otherwise $\pi_i > \pi_j$ if $i$ is after $j$.
For each pair of nodes, a weight constant $w_{ij}$ is given, and the objective function is to minimise $\Sigma_{ij} w_{ij} x_{ij}$, which has the effect of maximally fusing nodes, according to their weights.

@@

>>>>

\subsection{Haskell short-cut fusion}
Existing fusion systems for Haskell such as stream fusion\cite{coutts2007streamfusion, mainland2013haskell}, tend to cleverly reuse compiler optimisations such as inlining and rewrite rules\cite{jones2001playingby}, to fuse combinators without having to modify the compiler itself.
This approach has the advantage of simplicity, but is inherently limited in the amount of fusion it can perform.

@@

Talk about the simple formulation by Darte\cite{darte2002new}.
Has an integer variable for each node, denoting the number of the cluster it's in.
Also includes a binary variable for each node, which is whether the node is fused with all its successors --- in which case no array would be required, and an integer variable which is the maximum of all cluster numbers.
The objective function is to maximise the number of nodes that are completely fused, requiring no arrays, and minimise the maximum cluster number, which in turn minimises the number of clusters.
It doesn't require many constraints and is easy to implement, but doesn't work as well when there are loops of different sizes.
As loops of different sizes cannot be fused together, a simple method is to introduce an ordering on the sizes, and then extract loops of the same cluster number in order of size.
The problem here is that the objective function uses the maximum cluster number to minimise the number of loops, but this alone is no longer sufficient when there are multiple sizes.
\TODO{Explain why. Perhaps go into more detail about how arrays are contracted, which is different from Megiddo.}

The formulation by Megiddo\cite{megiddo1997optimal} supports different sized loops, and is therefore more relevant for our purposes.
For every pair of nodes $i,j$ in the LDG, a variable $x_{ij}$ is created, which denotes whether $i$ and $j$ are fused together.
Slightly awkwardly, but for simplicity of other constraints, $x_{ij} = 0$ if the two nodes are fused together.
If there is a fusion preventing edge between $i$ and $j$, then $x_{ij}$ is constrained to be $1$ --- that is, no fusion is possible.
This alone is not enough to guarantee a valid clustering.
To constrain the solution to acyclic and precedence preserving clusterings, a variable $\pi_i$ is added for each node $i$.
Constraints are added that require two nodes $i,j$ to have $\pi_i = \pi_j$ if $x_{ij} = 0$, and otherwise $\pi_i > \pi_j$ if $i$ is after $j$.
For each pair of nodes, a weight constant $w_{ij}$ is given, and the objective function is to minimise $\Sigma_{ij} w_{ij} x_{ij}$, which has the effect of maximally fusing nodes, according to their weights.

@@

\subsection{Non-optimal heuristics}

\TODO{Gao\cite{gao1993collective}} Collective loop fusion for array contraction.
I think this is a good introduction to the min/max edge following algorithm that is central to most of these.
Finds minimal number of clusters for single types. Despite the name, I don't think it actually minimises clustering for array contraction.

\CITEME{kennedy1993typed} Ken Kennedy --- Typed Fusion with Applications to Parallel and Sequential Code Generation.
Typed fusion --- choose an ordering of types, then find clustering of first type, fuse those together, then cluster second type, and so on.
It ain't optimal, even if you try all orderings (see Darte below)

\TODO{Chatterjee\cite{chatterjee1991size}}
Not optimal, heuristic based, etc, but does actually use ILP to reduce array crossing clusterings.
Like typed fusion, they choose clustering for different types rather arbitrarily, which is sub-par.

\CITEME{darte1998fusion} Alain Darte --- On the complexity of loop fusion.
Shows that loop fusion, minimising certain things (ie manifest arrays and number of loops) is NP-hard.

\TODO{Song\cite{song2004improving}} Improving data locality by array contraction.
All about \emph{controlled SFC}, ie shifting, fusion and contraction.
Formalises cost of memory references depending on distance / how many other references in between. Then only fuses loops if it doesn't raise the distance too much.

\CITEME{ashby2006iterative} T.J. Ashby --- Iterative collective loop fusion.
Executes the programs to decide which clustering is best, apparently. I didn't read it thoroughly, but didn't understand where they actually get the test data from. Unless it's a kind of JIT thing.
I honestly don't think much of it, but it \emph{is} more recent than the other stuff, so I suspect it'd be good to mention just so we're not ignoring ``modern fusion''.

\CITEME{kennedy2000fastgreedy} Ken Kennedy --- Fast Greedy Weighted Fusion.

\TODO{Sarkar\cite{sarkar1991optimization}} Optimization of array accesses by collective loop transformations.
This is probably not at all relevant, but it's \emph{very cute}. Use a two-colouring algorithm to decide when to reverse loops, to get best fusion. Not our scene.

\TODO{I don't really have a story on polyhedral analysis or unimodular.}
Now, here's my understanding of unimodular (and I'm just guessing this applies to polyhedral too):
if the dependency matrix for a loop nest or set of loops forms a \emph{unimodular matrix} (an integral matrix whose inverse is also integral), only then can it be dealt with by unimodular transforms.
